Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-27 paper/paper.tex
Updated Implementation Section

\section{Generating Liveness Equations}
\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
\end{align*}

@@

\begin{align*}
\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \cup \bX\sigma_{\mathit{\length}} \\
\Lanv{7}{\pu} &= \Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \\
\Lanv{7}{\py} &= \bX\sigma_{\mathit{\length}} \\
\Lanv{3}{\xl} &= \bX\sigma_{\mathit{\length}} \\
\Lanv{3}{\px} &= \bX\sigma_{\mathit{\length}}
\\[1ex]
  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
\end{align*}

@@

\section{Implementation}
% GC assumptions
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.
% GC points - mention reduction of gc points due to e-paths
An important  difference in GC of  lazy language compared  to an eager
language  is  the  point  of  garbage collection.  In  case  of  eager
languages only \CONS~ points and  function calls are the GC points. In
a lazy language  every \LET~ statement is a  potential GC point. Hence
we  need  to store  liveness  automata  corresponding  to every  \LET~
statement in the program. This leads  to an explosion in the number of
automata. From our observation that  along an {\em E-path} all program
points have  the same liveness, we  can reduce the  number of automata
state required  by sharing  the same automata  for all  program points
along an {\em E-path}. In the  example program, we only need to create
automata for program  points \ldots, since they are  the only distinct
{\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

% Liveness for cons cells & reachability for closures
During GC, if the heap cell being processed is a cons cell, we consult
the  liveness automaton  corresponding to  it and  copy only  the live
paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
approximate the  liveness to be {\{0+1\}\*} and  copy everything under
the closure.
% Handling of mixture of LGC & RGC
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closure or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a later GC
% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
% Proof for the correctness. 
% Difference between eager and lazy GC

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,

>>>>

\section{Generating Liveness Equations}
\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  (\CAR~\pa) \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pa) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pa.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{length}}}
\end{align*}

@@

\begin{align*}
\Lanv{7}{\xl} &= \acdr\Lf{length}{1}{\bX\sigma_{\mathit{length}}} \cup \bX\sigma_{\mathit{length}} \\
\Lanv{7}{\pu} &= \Lf{length}{1}{\bX\sigma_{\mathit{\length}}} \\
\Lanv{7}{\py} &= \bX\sigma_{\mathit{length}} \\
\Lanv{3}{\xl} &= \bX\sigma_{\mathit{length}} \\
\Lanv{3}{\px} &= \bX\sigma_{\mathit{length}}
\\[1ex]
  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
\end{align*}

@@

\section{Implementation}
% GC assumptions
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.
\subsection{GC points and automata sharing}
% GC points - mention reduction of gc points due to e-paths
We describe a simple outline of the GC algorthm in figure {\em add refernce to algorithm}.
In a lazy language each \LET~ expression creates a closure, and hence allocates memory 
on the heap. Therefore in contrast to an eager language, every \LET~ statement is a  potential GC point. 
As a GC can be triggered at any of the \LET~ expressions, we  need  to store  liveness  automata  
corresponding  to every  \LET~ statement in the program. 
Even though this leads  to an explosion in the number of
automata, we can optimize it using our observation that  along an {\em E-path} all program
points have  the same liveness. We therefore create a single liveness environment corresponding to each 
unique {\em E-path} sets and share the same automata  for all  program points
having a common {\em E-path} set. In the  example program {\em refer to the example program}, we only need to create
automata for the {\em E-path} sets \ldots, since they are  the only distinct
{\em E-path} sets in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

\subsection{Liveness based GC algorithm}
% Liveness for cons cells & reachability for closures
When ever the program runs out of memory, we scan the current buffer and we check if 
each reference is live or not and using a depth first traversal copy its children recursively.
The handling of a heap cell depends on the type of data it stores, as described in the algorithm{\em add reference to algorithm}.
\begin{itemize}
\item An atomic value is copied if it is live.
\item A \CONS~ cell is copied if it is live. Its \CAR~ and \CDR~ part are copied recursively if they are live.
\item A closure and everything reachable from it are copied if it is live otherwise nothing is copied. 
\end{itemize}

Observe that we are using liveness based GC for \CONS~ cells whereas we do a reachability based GC for closures. 
Since we do not nullify references that are not live this can lead to copying of incorrect references. 
Consider a \CONS~ cell {\em c} whose \CAR~ part is live but \CDR~ part is not live. During GC the pointer corresponding to the 
\CDR~ part is neither copied and nor nullified. Post GC if this cell becomes part of a closure then during a subsequent GC we will do 
a reachability based GC on {\em c} and hence try to chase the pointer corresponding to its \CDR~ part. 
One way to avoid this problem is by setting references which are not live to null. Doing this during GC is expensive as it 
requires an alias analysis. We avoid this overhead by waiting for the GC to end and then setting any references which point to 
the dead buffer to null. From the definition of liveness any reference which is declared not live will not be referenced beyond that 
execution point till the end of the execution. Hence once the GC is complete any pointer in the live buffer which points to a location 
in the dead buffer can be nullified safely. Although this involves scanning the entire live buffer again, the overhead is not too bad as 
the expected number of live heap cells in the live buffer (immediately after a GC) is usually much less compared to the size of the live buffer. Also, by 
copying closures using reachability we can perform certain optimizations which will not be possible if we use only liveness  based GC.
We describe this optimization in section{\em add reference}.
% A reference copied using LGC could become part of a closure in a later GC
% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
% Proof for the correctness. 
% Difference between eager and lazy GC

\subsection{Printing and GC on the print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,
